Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-12 analysis/paper/paper.Rmd
updates for discussion section

To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation for sets of data coupled with statistical comparison to identify lower value as an indicator of higher degree of standardization [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be assessed by comparing coefficient of variation and the differences between artisans can be detected at both intra-individual level and inter-individual level in a region. Although this traditional typological and linear measurements are useful for identifying differences in archaeological assembles, it might be limited because dimensional measurements can be insensitive to subtle variations resulting from changes in assemblages. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Using geometric morphomerics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts. The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in the indigenous society.

@@

The upper component experienced foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since the 19th century. Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were trade goods commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence. In addition to artifacts, features such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site [@Chen2007]. The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site. More than 550 thousands of sherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed. It is interesting to note that there are only two forms of vessels recovered, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between. Those pots show high consistency in shape. They are globular body with short neck and wide mouth. The exterior surface below the neck is treated with a wide variety of impressed geometric motifs. They were believed used for cooking because of the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior. Moreover, there is no other utilitarian earthenware ceramics found at this site suggests this type of pot was mainly used for daily cooking. Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.  
We examine 291 pots recovered from Kiwulan site. The layer from 1 to 6 could be divided into 3 time periods. Layer 5 and 6 represent pre-contact period, layer 4 represents contact period, and layer1, 2, and 3 represent post-contact period.  The amount of pots for each layer shows below. Although most pots are not complete, the thickness from rim, neck, to body can be measured. I have also measured the diameter of the rim, neck, and body. For those pots that are incomplete, the diameter is measured by its curvature. Since the height is incomplete for most pots, this preliminary analysis focuses more on the possible change in thickness and diameter of pot for different parts, and their ratio over time.

Petrographic analysis for thin sections was conducted at the Department of Geosciences, National Taiwan University. The results show high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate. The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%). The size of the particles ranges from 500 micron to 1300 micron (=1 mm). Overall, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz. Those minerals are believed to be local materials of the Yilan Plain.

@@

2. How does pottery change in shape in relation to foreign influences?
3. Did pottery become more homogeneous and standardized in shape after foreign contacts with the European colonizer or the Chinese immigrants?

@@

Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments specifically, after the presence of the European that hints the emergence of social inequality. We examine the degree of the ceramic standardization to measure ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century. There are significant differences between pre-European period and post-European period (P-value < 0.05), and between Chinese contact and either pre-European or post-European (P-value < 0.01). The average shape presents round body with wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period. In general, pottery shapes become shorter in hight over time that leads to oval-shaped body. The clay paste remains similar across different phases in relation to the increasing standardization of the pottery shape. More standardized shape in the Chinese period might indicate the appearance of specialized group.  To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric. However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric where not enough room for absorbing food residue [citation].

Although there is differences in pottery shape between pre-European and post-European period, it is more significant in the Chinese period with more standardized pottery shape. The results support part of our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society, however, stronger in the Chinese period instead of the European period that we originally thought to be. The hypothesis of emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported. The nature and context of the interaction with the foreign presences might give an insight into the different degree of standardized pottery. Compared to other region in Taiwan, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as trading base. Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [@Acabado2017]. On the contrary, the interaction between indigenous people and the Chinese in the 19th century was more intense and direct since those Chinese immigrants settled in Yilan and lived closely with indigenous societies. This direct influence is reflected by the archaeological evidence of large amount of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Also, the finding of adoption of coffin in mortuary practice that viewed as an ethnic symbol for Chinese was observed [@Chen2007].

Other than appearance of specialized groups, more standardized form without changes in clay paste suggests a preferred shape for pottery production. Another explanation could be related to an act of resistance to the Chinese immigrant by emphasizing the transitional practice, especially shape is a visible trait. @Raux2015

>>>>

To identify craft specialization, the variation of standardization of specialized product is commonly used as indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied in archaeological assemblages by calculating coefficient of variation for sets of data coupled with statistical comparison to identify lower value as an indicator of higher degree of standardization [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be assessed by comparing coefficient of variation and the differences between artisans can be detected at both intra-individual level and inter-individual level in a region. Although this traditional typological and linear measurements are useful for identifying differences in archaeological assembles, it might be limited because dimensional measurements can be insensitive to subtle variations resulting from changes in assemblages. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Using geometric morphometrics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts. The aim of this paper is to investigate if there are increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in the indigenous society.

@@

The upper component experienced foreign contacts including the European colonial presence in the 17th century and great waves of Chinese immigrants since the 19th century. Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were trade goods commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence. In addition to artifacts, features such as burials, middens, and post-holes with in-situ posts that explain Kiwulan was a large settlement site [@Chen2007]. The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site. More than 550 thousands of sherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed. It is interesting to note that there are only two forms of vessels recovered, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between. Those pots show high consistency in shape. They are globular body with short neck and wide mouth. The exterior surface below the neck is treated with a wide variety of impressed geometric motifs. They were believed used for cooking because of the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior. Moreover, there is no other utilitarian earthenware ceramics found at this site suggests this type of pot was mainly used for daily cooking. Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand.

We examine 291 pots recovered from Kiwulan site. The layer from 1 to 6 could be divided into 3 time periods. Layer 5 and 6 represent pre-contact period, layer 4 represents contact period, and layer1, 2, and 3 represent post-contact period.  The amount of pots for each layer shows below. Although most pots are not complete, the thickness from rim, neck, to body can be measured. I have also measured the diameter of the rim, neck, and body. For those pots that are incomplete, the diameter is measured by its curvature. Since the height is incomplete for most pots, this preliminary analysis focuses more on the possible change in thickness and diameter of pot for different parts, and their ratio over time.

Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate. The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%). The size of the particles ranges from 500 micron to 1300 micron (=1 mm). In general, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz. The composition matches the mineralogical composition of local raw materials in the Yilan Plain [\citation].

@@

@@

Previous investigations at Kiwulan suggest an unequal distribution pattern of prestige goods, trade ornaments specifically, after the presence of the European that hints the emergence of social inequality. We examine the degree of the ceramic standardization to measure ceramic specialization that is usually associate with social differentiation, such as the elite control [@Costin2001; @Junker1999]. The results confirm that the differences in pottery shapes can be detected using EFA method, indicating the foreign influences on the indigenous society- the Europeans in the 17th century and the Chinese in the 19th century. The result of MANOVA shows there is significant difference in shapes between pre-European and Post-European period, and between the pre-European and the Chinese period. The average shape presents round body with wider rim and neck before the European contact, and turns to oval-shaped body with narrower rim and neck after the European presence and even more obvious during the Chinese contact period. In general, pottery shapes become shorter in hight over time that leads to oval-shaped body. For the standardization of pottery shape, the result of CV test indicates the significant differences between pre-European period and post-European period (P-value < 0.05), and between Chinese contact and either pre-European or post-European (P-value < 0.01). This suggests pottery shape becomes more homogeneous and standardized after foreign contacts with the European colonizer and even more with the Chinese immigrants. The compositional analysis shows clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting the raw material sources remains unchanged. More standardized shape in the Chinese period might indicate the appearance of specialized group. To figure out whether this change is related to the function of pot, we made an attempt to conduct lipid analysis for potsherds samples to identify the source of residue absorbed into the fabric. However, we did not obtain useful results due to a low lipid yield, which might result from the thin and less porous fabric where not enough room for absorbing food residue [citation].

The results support part of our hypothesis that foreign influences had impacts on the emergence of social inequality in local indigenous society reflected by the degree of pottery standardization, however, stronger in the Chinese period instead of the European period originally thought to be. The hypothesis of emergence of social inequality based on the monopolies for long-distance trade brought by the Europeans was not fully supported. The nature and context of the interaction with the foreign presences might give an insight into the different degree of standardized pottery. Compared to other region in Taiwan, the European colonial control was weak in Yilan due to the relatively isolated location by mountains and the economic consideration of the Spanish and the Dutch who prefer northern Taiwan as trading base. Yilan could be seen as a place experienced indirect influences by the European trade network and their colonial activities in a pericolonial context [@Acabado2017]. On the contrary, the interaction between indigenous people and the Chinese in the 19th century was more intense and direct since those Chinese immigrants settled in Yilan and lived closely with indigenous societies. This direct influence is reflected by the archaeological evidence of large amount of Chinese porcelains and distinctive architectural bricks and tiles used by Chinese [@Hsieh2009]. Also, the finding of adoption of coffin in mortuary practice that viewed as an ethnic symbol for Chinese was observed [@Chen2007].

More standardized pottery shape without controlling the clay paste or changes in technology suggests the shape as a preferred characteristic in pottery production. In addition to the appearance of specialized groups, another explanation could be related to an act of resistance to the Chinese immigrant by emphasizing the traditional pottery production practice, of which shape is a visible trait for. @Raux2015
